---
title: "Higher Education in 2018-2019"
author: Kelsey Pearson
output: html_document
date: "2023-09-18"
---

```{r setup, message=FALSE, warning=FALSE, echo=F}
knitr::opts_chunk$set(echo = TRUE)

library(pacman)
p_load(tidyr, readr, ggplot2, dplyr, lubridate, 
       readxl, stringr, janitor, kbl, kableExtra, DT)
```

To demonstrate some of the essential skills required for this role, candidates are asked to produce a piece of work for presentation at the interview. 

Using data freely available via The Higher Education Statistics Agency (HESA) at https://www.hesa.ac.uk/data-and-analysis, for 2018/19 academic year, produce a report showing how the University of Edinburgh compares with the University of Aberdeen, the University of Glasgow, the University of Strathclyde and the University of St Andrews on the following metrics:

* Number of Undergraduate and Postgraduate Taught students
* The ratio of students to staff (‘SSR’)
* Income to University per academic staff member

You should clearly note any assumptions you have made as part of the analysis. 

## Number of Undergraduate and Postgraduate Taught students

```{r load-student-data, message=FALSE, warning=FALSE, echo=F}
students = read_csv('data/student-ug-pg-(2018-19).csv', skip = 14, 
                    show_col_types = FALSE) %>% 
  clean_names()
```

Data sourced from: https://www.hesa.ac.uk/data-and-analysis/students/table-1

Method:

Filter for the following universities: University of Edinburgh; University of Aberdeen, University of Glasgow; University of Strathclyde; and University of St. Andrews

Filter for following levels of study: All undergraduate; and Postgraduate (taught)

Include all years of study, mode of study, and do not consider other categories such as sex or domicile

```{r distinct-categories, message=FALSE, warning=FALSE, echo=F}
# Identify filters
distinct_he = students %>% 
  distinct(he_provider)

distinct_los = students %>% 
  distinct(level_of_study)

distinct_fym = students %>% 
  distinct(first_year_marker)

distinct_mos = students %>% 
  distinct(mode_of_study)
```

```{r filter-unis, message=FALSE, warning=FALSE, echo=F}
he_filter = c('The University of Aberdeen', 'The University of Edinburgh',
              'The University of Glasgow', 'The University of Strathclyde',
              'The University of St. Andrews')

los_filter = c('Postgraduate (taught)', 'All undergraduate')

# Filter dataset
ug_pg_students = students %>% 
  filter(he_provider %in% he_filter) %>% 
  filter(level_of_study %in% los_filter) %>% 
  filter(first_year_marker == "All") %>% 
  filter(mode_of_study == "All") %>% 
  filter(category == "Total") %>% 
  filter(country_of_he_provider == "Scotland" & region_of_he_provider == "Scotland") %>%
  mutate(total= sum(number), .by = he_provider)
```

```{r plot-students, message=FALSE, echo=F}
# Plot
ug_pg_students %>% 
  ggplot(aes(x = reorder(he_provider, total), y = number, 
             group = level_of_study, fill = level_of_study)) + 
  geom_bar(stat = 'identity', position = position_dodge2(reverse = T)) +
  geom_text(aes(label = format(number, big.mark = ','),
                x = reorder(he_provider, total)),
            hjust = 0, colour = "black", 
            position = position_dodge2(.9, reverse = T),
            size = 3) +
  labs(title = 'Number of students in 2018-2019',
       x = '',
       y = '') +
  scale_x_discrete(labels = function(x) str_wrap(str_replace_all(x, "foo" , " "),
                                                 width = 15)) +
  scale_y_continuous(limits = c(NA, 25000)) +
  coord_flip() +
  theme_minimal() +
  theme(legend.position = 'bottom',
        legend.title = element_blank(),
        plot.margin = margin(r = 60))
```

```{r, message=FALSE, warning=FALSE, echo=F}
# Table
table_ug_pg_students = ug_pg_students %>% 
  select(he_provider,
         level_of_study,
         number) %>% 
  pivot_wider(names_from = level_of_study, values_from = number) %>% 
  mutate(total = rowSums(across(where(is.numeric)))) %>% 
  arrange(desc(total))

table_ug_pg_students %>% 
  mutate(`Postgraduate (taught)` = format(`Postgraduate (taught)`, big.mark = ","),
         `All undergraduate` = format(`All undergraduate`, big.mark = ","),
         total = format(total, big.mark = ",")) %>% 
  kbl(col.names = c("HE Provider", "Postgraduate (taught)", "Undergraduate", "Total"),
      caption = "Student population in 2018-2019 (sorted by Total)")
```

`r table_ug_pg_students %>% filter(total == max(total)) %>% pull(he_provider)` has the largest student population with `r format(table_ug_pg_students %>% filter(total == max(total)) %>% pull(total), big.mark = ",")` students studying at Undergraduate level (`r format(table_ug_pg_students %>% filter(total == max(total)) %>% pull('All undergraduate'), big.mark = ",")`) and in Postgraduate Taught coures (`r format(table_ug_pg_students %>% filter(total == max(total)) %>% pull('Postgraduate (taught)'), big.mark = ",")`).

`r table_ug_pg_students %>% filter(total == min(total)) %>% pull(he_provider)` has the smallest student population with `r format(table_ug_pg_students %>% filter(total == min(total)) %>% pull(total), big.mark = ",")` students studying in Undergraduate level (`r format(table_ug_pg_students %>% filter(total == min(total)) %>% pull('All undergraduate'), big.mark = ",")`) and in Postgraduate Taught courses (`r format(table_ug_pg_students %>% filter(total == min(total)) %>% pull('Postgraduate (taught)'), big.mark = ",")`).

## Ratio of students to staff ('SSR')

Number of students per member of staff. Need to consider FTE, placements

Assumptions and considerations:

Undergraduate and postgraduate taught students in full-time education

Academic staff, full time, non-atypical contract

SSR calculation

```{r load-staff-data, message=FALSE, warning=FALSE, echo=F}
staff = read_csv('data/staff.csv', skip = 33, 
                    show_col_types = FALSE) %>% 
  clean_names()
```

```{r distinct-staff-categories, message=FALSE, warning=FALSE, echo=F}

# Identify filters
distinct_co_m = staff %>% 
  distinct(contract_marker)

distinct_ac_m = staff %>% 
  distinct(academic_marker)

distinct_at_m = staff %>% 
  distinct(atypical_marker)

distinct_moe = staff %>% 
  distinct(mode_of_employment)

distinct_activity = staff %>% 
  distinct(activity_standard_occupational_classification)
```

```{r filter-staff, message=FALSE, warning=FALSE, echo=F}
he_filter = c('The University of Aberdeen', 'The University of Edinburgh',
              'The University of Glasgow', 'The University of Strathclyde',
              'The University of St. Andrews')

los_filter = c('Postgraduate (taught)', 'All undergraduate')

# Filter dataset
he_staff = staff %>% 
  filter(academic_year == "2018/19") %>% 
  filter(he_provider %in% he_filter) %>% 
  filter(country_of_he_provider == "Scotland" & region_of_he_provider == "Scotland")
```

```{r all-staff, message=FALSE, warning=FALSE, echo=F}
table_staff = he_staff %>% 
  filter(mode_of_employment == "All" & academic_marker == "All") %>% 
  summarise(total_staff = sum(number), .by = he_provider)

table_staff %>% 
  arrange(desc(total_staff)) %>% 
  kbl(col.names = c("HE Provider", "Total"),
      caption = "Staff levels in 2018-2019")
```

```{r ssr-staff, message=FALSE, warning=FALSE, echo=F}
# SSR for all students and all staff
table_ssr = table_ug_pg_students %>% 
  left_join(table_staff, by = join_by(he_provider == he_provider)) %>% 
  mutate(ssr = round(total / total_staff, digits = 2)) %>% 
  select(he_provider, total, total_staff, ssr)

table_ssr %>% 
  arrange(ssr) %>% 
  kbl(col.names = c("HE Provider", "Students", "Total Staff", "SSR"),
      caption = "SSR when considering Undergraduate and Postgraduate taught students and all staff in 2018-2019")
```

```{r all_academic-staff, message=FALSE, warning=FALSE, echo=FALSE}
# Student:Academic Staff Ratio
table_academic_staff = he_staff %>% 
  filter(activity_standard_occupational_classification == "Total academic staff"
         | activity_standard_occupational_classification != "Total academic atypical staff") %>% 
  filter(mode_of_employment == "Full-time") %>% 
  distinct(number, he_provider, .keep_all = T) %>% 
  select(he_provider, activity_standard_occupational_classification, number) %>% 
  summarise(total_academic_staff = sum(number), .by = he_provider) %>% 
  arrange(desc(total_academic_staff))

table_sasr = table_ug_pg_students %>% 
  left_join(table_academic_staff, by = join_by(he_provider == he_provider)) %>% 
  mutate(ssr = round(total / total_academic_staff, digits = 2)) %>% 
  select(he_provider, total, total_academic_staff, ssr)

table_sasr %>% 
  arrange(ssr) %>% 
  kbl(col.names = c("HE Provider", "Students", "Academic staff", "SSR"),
      caption = "SSR when considering Undergraduate and Postgraduate taught students in full-time education and full-time academic staff in 2018-2019")
```

`r table_sasr %>% filter(ssr == min(ssr)) %>% pull(he_provider)` has the smallest student to staff ratio, with an SSR of `r format(table_sasr %>% filter(ssr == min(ssr)) %>% pull(ssr))`. `r table_sasr %>% filter(ssr == max(ssr)) %>% pull(he_provider)` has the smallest student to staff ratio, with an SSR of `r format(table_sasr %>% filter(ssr == max(ssr)) %>% pull(ssr))`.

## Income to University per academic staff member

```{r load-finance, message=FALSE, warning=FALSE, echo=F}

finance = read_csv('data/finances.csv', skip = 12, 
                    show_col_types = FALSE) %>% 
  clean_names() %>% 
  mutate(value_000s = as.numeric(value_000s))
```

```{r filter-finance, message=FALSE, warning=FALSE, echo=F}

# Filter dataset
table_finance = finance %>% 
  filter(academic_year == "2018/19") %>% 
  filter(he_provider %in% he_filter) %>% 
  filter(country_of_he_provider == "Scotland"
         & region_of_he_provider == "Scotland") %>%
  filter(category_marker == "Income" 
         & category == "Total income" 
         & year_end_month == "All") %>% 
  select(he_provider,
         total_income = value_000s) %>% 
  mutate(total_income_curr = format(total_income, big.mark = ","))
```

```{r finance-points, message=FALSE, warning=FALSE, echo=F}
max_income_uni = table_finance %>% 
  filter(total_income == max(total_income)) %>%
  pull(he_provider)

max_income = table_finance %>% 
  filter(total_income == max(total_income)) %>% 
  pull(total_income)

min_income_uni = table_finance %>% 
  filter(total_income == min(total_income)) %>%
  pull(he_provider)

min_income = table_finance %>% 
  filter(total_income == min(total_income)) %>% 
  pull(total_income)
```

The `r max_income_uni` had the largest income in 2018-2019, with an income of `r paste0("£", format(max_income, big.mark = ","))`.

In the same year, the r min_income_uni had the smallest income with `r paste0("£", format(min_income, big.mark = ","))`. The `r paste0(max_income_uni, "'s")` income is `r round(max_income / min_income, digits = 1)` times greater than the `r min_income_uni`.

```{r income-per-staff, message=FALSE, warning=FALSE, echo=F}

table_income_per_staff = table_academic_staff %>% 
  left_join(table_finance, by = join_by(he_provider == he_provider)) %>% 
  mutate(income_per_staff = round(total_income / total_academic_staff, digits = 2)) %>% 
  arrange(desc(income_per_staff))

table_income_per_staff %>% 
  select(-total_income) %>% 
  mutate(total_academic_staff = format(total_academic_staff, big.mark = ","),
         total_income_curr = paste0("£", total_income_curr),
         income_per_staff = paste0("£", income_per_staff)) %>% 
  kbl(col.names = c("HE Provider", "Academic staff", 
                    "Total income (£)", "Income per\nacademic staff member (£)"),
      caption = "Income per academic staff member in 2018-2019")
```

```{r test, eval=FALSE, message=FALSE, warning=FALSE, echo=F}
# the last two columns are Pound Sterling, and round column E to 3 decimal places
datatable(table_income_per_staff, 
          colnames = c("HE Provider", "Total academic staff",
                       "Total income",
                       "Income per academic staff member")) %>% 
  formatCurrency(3:4, currency = "£") 
```

