---
title: "Higher Education in 2018-2019"
output: html_document
date: "2023-09-18"
---

```{r setup, message=FALSE, warning=FALSE, echo=F}
knitr::opts_chunk$set(echo = TRUE)

install.packages(setdiff("pacman", rownames(installed.packages())))
library("pacman")
p_load(tidyr, readr, ggplot2, dplyr, lubridate, 
       readxl, stringr,janitor, kbl, kableExtra, DT)
```

## Business Intelligence Analyst

###Pre-interview preparation

To demonstrate some of the essential skills required for this role, candidates are asked to produce a piece of work for presentation at the interview. 

Using data freely available via The Higher Education Statistics Agency (HESA) at https://www.hesa.ac.uk/data-and-analysis, for 2018/19 academic year, produce a report showing how the University of Edinburgh compares with the University of Aberdeen, the University of Glasgow, the University of Strathclyde and the University of St Andrews on the following metrics:

* Number of Undergraduate and Postgraduate Taught students
* The ratio of students to staff (‘SSR’)
* Income to University per academic staff member

You should clearly note any assumptions you have made as part of the analysis. 

## Number of Undergraduate and Postgraduate Taught students

```{r load-student-data, message=FALSE, warning=FALSE, echo=F}

students = read_csv('data/student-ug-pg-(2018-19).csv', skip = 14, 
                    show_col_types = FALSE) %>% 
  clean_names()
```

Data sourced from: https://www.hesa.ac.uk/data-and-analysis/students/table-1

Method:

Filter for the following universities: University of Edinburgh; University of Aberdeen, University of Glasgow; University of Strathclyde; and University of St. Andrews

Filter for following levels of study: All undergraduate; and Postgraduate (taught)

Include all years of study, mode of study, and do not consider other categories such as sex or domicile

```{r distinct-categories, message=FALSE, warning=FALSE, echo=F}

# Identify filters
distinct_he = students %>% 
  distinct(he_provider)

distinct_los = students %>% 
  distinct(level_of_study)

distinct_fym = students %>% 
  distinct(first_year_marker)

distinct_mos = students %>% 
  distinct(mode_of_study)
```

```{r filter-unis, message=FALSE, warning=FALSE, echo=F}
he_filter = c('The University of Aberdeen', 'The University of Edinburgh',
              'The University of Glasgow', 'The University of Strathclyde',
              'The University of St. Andrews')

los_filter = c('Postgraduate (taught)', 'All undergraduate')

# Filter dataset
ug_pg_students = students %>% 
  filter(he_provider %in% he_filter) %>% 
  filter(level_of_study %in% los_filter) %>% 
  filter(first_year_marker == "All") %>% 
  filter(mode_of_study == "All") %>% 
  filter(category == "Total") %>% 
  filter(country_of_he_provider == "Scotland" & region_of_he_provider == "Scotland") %>%
  mutate(total= sum(number), .by = he_provider)
```

```{r plot-students, message=FALSE, echo=F}

ug_pg_students %>% 
  ggplot(aes(x = reorder(he_provider, total), y = number, 
             group = level_of_study, fill = level_of_study)) + 
  geom_bar(stat = 'identity', position = position_dodge2(reverse = T)) +
  geom_text(aes(label = format(number, big.mark = ','),
                x = reorder(he_provider, total)),
            hjust = 0, colour = "black", 
            position = position_dodge2(.9, reverse = T),
            size = 3) +
  labs(title = 'Number of students in 2018-2019',
       x = '',
       y = '') +
  scale_x_discrete(labels = function(x) str_wrap(str_replace_all(x, "foo" , " "),
                                                 width = 15)) +
  scale_y_continuous(limits = c(NA, 25000)) +
  coord_flip() +
  theme_minimal() +
  theme(legend.position = 'bottom',
        legend.title = element_blank(),
        plot.margin = margin(r = 60))
```

```{r, message=FALSE, warning=FALSE, echo=F}
table_ug_pg_students = ug_pg_students %>% 
  select(he_provider,
         level_of_study,
         number) %>% 
  pivot_wider(names_from = level_of_study, values_from = number) %>% 
  mutate(total = rowSums(across(where(is.numeric)))) %>% 
  arrange(desc(total))

table_ug_pg_students %>% 
  mutate(`Postgraduate (taught)` = format(`Postgraduate (taught)`, big.mark = ","),
         `All undergraduate` = format(`All undergraduate`, big.mark = ","),
         total = format(total, big.mark = ",")) %>% 
  kbl(col.names = c("HE Provider", "Postgraduate (taught)", "Undergraduate", "Total"),
      caption = "Student population in 2018-2019 (sorted by Total)")
```

`r table_ug_pg_students %>% filter(total == max(total)) %>% pull(he_provider)` has the largest student population with `r format(table_ug_pg_students %>% filter(total == max(total)) %>% pull(total), big.mark = ",")` students studying at Undergraduate level `(r format(table_ug_pg_students %>% filter(total == max(total)) %>% pull('All undergraduate'), big.mark = ","))` and in Postgraduate Taught coures `(r format(table_ug_pg_students %>% filter(total == max(total)) %>% pull('Postgraduate (taught)'), big.mark = ","))`.

`r table_ug_pg_students %>% filter(total == min(total)) %>% pull(he_provider)` has the smallest student population with `r format(table_ug_pg_students %>% filter(total == min(total)) %>% pull(total), big.mark = ",")` students studying in Undergraduate level `(r format(table_ug_pg_students %>% filter(total == min(total)) %>% pull('All undergraduate'), big.mark = ","))` and in `Postgraduate Taught courses (r format(table_ug_pg_students %>% filter(total == min(total)) %>% pull('Postgraduate (taught)'), big.mark = ","))`.

## Income to University per academic staff member

```{r load-finance, message=FALSE, warning=FALSE, echo=F}

finance = read_csv('data/finances.csv', skip = 12, 
                    show_col_types = FALSE) %>% 
  clean_names() %>% 
  mutate(value_000s = as.numeric(value_000s))
```

```{r filter-finance, message=FALSE, warning=FALSE, echo=F}

# Filter dataset
table_finance = finance %>% 
  filter(academic_year == "2018/19") %>% 
  filter(he_provider %in% he_filter) %>% 
  filter(country_of_he_provider == "Scotland"
         & region_of_he_provider == "Scotland") %>%
  filter(category_marker == "Income" 
         & category == "Total income" 
         & year_end_month == "All") %>% 
  select(he_provider,
         total_income = value_000s) %>% 
  mutate(total_income_curr = format(total_income, big.mark = ","))
```

```{r finance-points, message=FALSE, warning=FALSE, echo=F}
max_income_uni = table_finance %>% 
  filter(total_income == max(total_income)) %>%
  pull(he_provider)

max_income = table_finance %>% 
  filter(total_income == max(total_income)) %>% 
  pull(total_income)

min_income_uni = table_finance %>% 
  filter(total_income == min(total_income)) %>%
  pull(he_provider)

min_income = table_finance %>% 
  filter(total_income == min(total_income)) %>% 
  pull(total_income)
```

The `r max_income_uni` had the largest income in 2018-2019, with an income of `r paste0("£", format(max_income, big.mark = ","))`.

In the same year, the r min_income_uni had the smallest income with `r paste0("£", format(min_income, big.mark = ","))`. The `r paste0(max_income_uni, "'s")` income is `r round(max_income / min_income, digits = 1)` times greater than the r min_income_uni.

```{r income-per-staff, message=FALSE, warning=FALSE, echo=F}

table_income_per_staff = table_academic_staff %>% 
  left_join(table_finance, by = join_by(he_provider == he_provider)) %>% 
  mutate(income_per_staff = round(total_income / total_academic_staff, digits = 2)) %>% 
  arrange(desc(income_per_staff))

table_income_per_staff %>% 
  select(-total_income) %>% 
  mutate(total_academic_staff = format(total_academic_staff, big.mark = ","),
         total_income_curr = paste0("£", total_income_curr),
         income_per_staff = paste0("£", income_per_staff)) %>% 
  kbl(col.names = c("HE Provider", "Academic staff", 
                    "Total income (£)", "Income per\nacademic staff member (£)"),
      caption = "Income per academic staff member in 2018-2019")
```

```{r test, eval=FALSE, message=FALSE, warning=FALSE, echo=F}
# the last two columns are Pound Sterling, and round column E to 3 decimal places
datatable(table_income_per_staff, 
          colnames = c("HE Provider", "Total academic staff",
                       "Total income",
                       "Income per academic staff member")) %>% 
  formatCurrency(3:4, currency = "£") 
```

